Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2021-02-10 manuscript_Huang_rev.tex
Updates from Overleaf

Due to our choice of $\tau$ and $D$, all SCMs take values of zero for the period-3 series because the temporal distance between each pair of components of the embedding vector covers a multiple of two full periods (cf.~Fig.~\ref{fig:embed}). Other choices of $D$ and $\tau$ would commonly yield non-zero values for the period-3 regime. For the three other cases of band merging, laminar state and outer crisis, the values of all SCMs significantly differ from zero. The pattern frequency based SCM $\mathcal{C}_O$ shows zero value and $\mathcal{H}_{O} = 1$, which implies a positioning in the lower right corner of the complexity--entropy plane in Fig.~\ref{fig:CElogistic}(a). However, for the transition matrix based SCMs, both the estimated entropies $\mathcal{H}_W$, $\mathcal{H}_E$ and associated complexity measures $\mathcal{C}_W$, $\mathcal{C}_E$ show results that are more consistent with the expectations when considering the respective level of chaoticity of the system in those three regimes (i.e., a rising Lyapunov exponent with increasing $r$ in the respective chaotic regimes). In particular, the estimated SCMs take their largest values for $r = 4$ in the outer crisis regime, and the SCMs are the smallest when $r = 3.679$ in the band merging case, while the laminar regime displays intermediate SCM values.

\subsection{Complexity--entropy planes} \label{sec:plane}
\subsubsection{Logistic map} 
In order to put our previous results into a broader context, we next show the complexity--entropy planes obtained when varying the control parameter $r$ in the logistic map. This simple nonlinear system has already been widely discussed in the framework of SCMs \cite{RossoPRE2007,MartinPLA2003}, but only for pattern frequency based complexity measures. For each value of $r$, we construct an OPTN from a time series of length $L = 10 ^ 6$ with the embedding parameters $D = \tau \in \{2,\ldots,7\}$.

As already stated above, when only one unique ordinal pattern is identified in a given time series, all SCMs take a value of $0$, which is reasonable since only self-loops are observed. This however happens only if the embedding parameters are chosen such that they exactly coincide with the periodicity of a certain periodic window (see Fig.~\ref{fig:embed}(g)). 
%Another practical aspect concerns the data requirements for computing entropies and complexity measures. For a reliable estimation of SCMs, a pragmatic condition on the length $L$ of the time series to be sufficiently larger than $D!$ \cite{rossoPRL2007,kowalskiPhyD2007} has been satisfied in our setting for $D \leq 7$. In the particular example of the logistic map, one may easily generate much longer realizations, which however {\color{red} is subjected to finite size effects in experimental time series analysis}.

@@

Figure~\ref{fig:CElogistic} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar when $D$ and $\tau$ are varied in the range $\{3,\ldots, 7\}$. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. {\color{red}The curves of $\mathcal{C}_{max}$ for different embeddings have been further annotated in Fig. \ref{fig:CElogistic}. In contrast, the $\mathcal{C}_{min}$ values are close to each other, showing less dependence on $D$, and therefore they are not annotated. }
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{(Color online) Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a, d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b, e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c, f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The {\color{red} dotted lines correspond to $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ in dependence on the respective entropy for $D=4$, and the dashed lines are $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ for $D=5$. Panels (a-c) show the results for embedding dimensions $D = 4,5$ with annotations for the respective $\mathcal{C}_{max}$ only} and (d-f) for $D = 6$. In panels (a, d), the bottom-right rectangles of dotted red lines indicate the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. However, the lower left rectangles of (b, c, e, f) correspond to regions of lower complexities. }  \label{fig:CElogistic}}
\end{figure*}

>>>>

Due to our \textcolor{red}{specific} choice of $\tau$ and $D$, all SCMs take values of zero for the period-3 series because the temporal distance between each pair of components of the embedding vector covers a multiple of two full periods (cf.~Fig.~\ref{fig:embed}). Other choices of $D$ and $\tau$ would commonly yield non-zero values for the period-3 regime. For the three other cases of band merging, laminar state and outer crisis, the values of all SCMs significantly differ from zero. The pattern frequency based SCM $\mathcal{C}_O$ shows zero value and $\mathcal{H}_{O} = 1$, which implies a positioning in the lower right corner of the complexity--entropy plane in Fig.~\ref{fig:CElogistic}(a). However, for the transition matrix based SCMs, both the estimated entropies $\mathcal{H}_W$, $\mathcal{H}_E$ and associated complexity measures $\mathcal{C}_W$, $\mathcal{C}_E$ show results that are more consistent with the expectations when considering the respective level of chaoticity of the system in those three regimes (i.e., a rising Lyapunov exponent with increasing $r$ in the respective chaotic regimes). In particular, the estimated SCMs take their largest values for $r = 4$ in the outer crisis regime, and the SCMs are the smallest when $r = 3.679$ in the band merging case, while the laminar regime displays intermediate SCM values.

\subsection{Complexity--entropy planes} \label{sec:plane}
\subsubsection{Logistic map} 
In order to put our previous results into a broader context, we next show the complexity--entropy planes obtained when varying the control parameter $r$ in the logistic map. This simple nonlinear system has already been widely discussed in the framework of SCMs \cite{RossoPRE2007,MartinPLA2003}, but only for pattern frequency based complexity measures. For each value of $r$, we \textcolor{red}{again} construct OPTN\textcolor{red}{s} from time series of length $L = 10 ^ 6$ with the embedding parameters $D = \tau \in \{2,\ldots,7\}$. {\color{red}As far as the classical ordinal pattern frequency based SCM is concerned, the given time series length is compatible with the chosen range of $D$ values, while the largest embedding dimension of $7$ is at the edge of the range where the ``optimistic'' estimate in Sec.~\ref{sec:tslength} can be considered to apply for transition frequency based SCMs as well.

In the context of finite sample size effects, we note that for the particular example of the logistic map, one could easily generate realizations of much longer lengths. Our numerical results show, however, that longer realizations of a length of up to $L\leq5\cdot 10^6$ do not change the obtained entropy and SCM \emph{estimates} significantly (not shown). From a practical perspective, two aspects call for attention in interpreting the obtained entropy and SCM values quantitatively. On the one hand, due to the finite numerical precision of the computed trajectories, any chaotic regime may in fact turn out to appear as periodic (with arbitrarily long periods); this can lead to numerical artifacts that will be further discussed in Sec.~\ref{sec:transi}. On the other hand, there may exist asymptotic analytical results for the case of $L\to\infty$ and $D\to\infty$, e.g., for the case of $r=4$ where one would expect $\mathcal{H}_O\to 1$ and $\mathcal{C}_O\to 0$, which substantially differ from our numerical results. In this regard, all estimated values of the different entropy and complexity measures always need to be considered as finite sample estimates that may well deviate from their expected asymptotic values and thereby depend on the specific values of $L$ and $D$ available for the analysis. This is, however, a common feature of other existing symbolic dynamics based entropies and SCMs.

As another introductory side note, we recall that as} already stated above, when only one unique ordinal pattern is identified in a given time series, all SCMs take a value of $0$, which is reasonable since only self-loops are observed. This however happens only if the embedding parameters are chosen such that they exactly coincide with the periodicity of a certain periodic window (see Fig.~\ref{fig:embed}(g)). \textcolor{red}{Hence, it cannot be expected that a given fixed choice of the embedding parameters will be able to highlight all periodic windows of arbitrary order in exactly the same quantitative way.}

@@

Figure~\ref{fig:CElogistic} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar \textcolor{red}{for $D=\tau=3,\ 7$ as well (not shown)}. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. {\color{red}The curves of $\mathcal{C}_{max}$ for different embeddings have been further annotated in Fig. \ref{fig:CElogistic}. In contrast, the $\mathcal{C}_{min}$ values are close to each other, showing less dependence on $D$, and therefore they are not annotated. }
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{(Color online) Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a, d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b, e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c, f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The {\color{red} dotted lines correspond to $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ in dependence on the respective entropy for $D=4$, and the dashed lines are $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ for $D=5$. Panels (a-c) show the results for embedding dimensions $D = 4,5$ with annotations for the respective $\mathcal{C}_{max}$ only} and (d-f) for $D = 6$. In panels (a, d), the bottom-right rectangles of dotted red lines indicate the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. However, the lower left rectangles of (b, c, e, f) correspond to regions of lower complexities. }  \label{fig:CElogistic}}
\end{figure*}
